### 2018-04-16 contents/reviews/multiplesclerosis.tex
Updates from ShareLaTeX

\section{Multiple Sclerosis}
This chapter will review multiple sclerosis (MS) as a disorder, given the importance of MS-related TN as an independent TN subtype. MS is a chronic autoimmune demyelination disease of the central nervous system. It is a progressive disease characterized by wide-spread lesion, or plaques, within the brain, spinal cord, and optic nerve. Its pathogenesis is related to the focal inflammation and damage of the central myelin, where the disruptions of the myelin sheath negatively affect signal transduction of the central white-matter, and thereby result in severe disability. MS symptoms can include cognitive deficits, fatigue, depression, motor weakness affecting the upper and lower extremities and neuropathic pain. Currently there is no cure for MS. An estimated 2,300,000 people world wide are affected by MS, and 50\% of the patient will require motor assistance 15 years after disease onset \cite{Goldenberg2012}. The course of MS can be highly variable due to the distinct plaque distribution pattern in each individual, and therefor MS is difficult to diagnose and treat.

@@

Positive MS diagnosis is satisfied by A) At least two different lesions are present in the CNS white matter (space dissemination); B) At least two different episodes must occur in the disease course (time dissemination); and C) Chronic inflammation of the CNS. One or more of these criteria result in a general diagnosis of MS. Both space and time dissemination criteria are confirmed with clinical MRI. Detailed disability ratings are confirmed by the Expanded Disability Status Scale (EDSS)\cite{Kurtzke1983}, which tracks the progression of MS symptoms in an individual.

MS is defined by the existence of focal demyelination in the CNS that manifests as lesions or plaques. Ultrastructural lesion analysis shows demyelination in the form of myelin debris, macrophages, lymphocytes and Creutzfeldt-Peters cells. Partial remyelination by oligodendrocytes can be present, as well as signs of axonal swelling and injury. In tissue samples of longstanding MS there is no presence of acute demyelination, and heavy loss of axons and myelin could be observed \cite{Filippi2012}.

@@

The MS lesions or plaques are characterized as focal hyperintensities in T2-weighted or proton-density images. Chronic and acute lesions could be distinguished by the presence of blood-brain-barrier (BBB) breakdown with the use of MR contrast agents such as gadolinium. Focal MS lesions commonly appear as round hyperintensities in MR images ranging from a few milimeters to centimeter in size. They are also found densely dispersed in paraventricular white matter. The presence of MR hyperintensities is not unique to MS, therefore the location, shape and time dissemination must be considered for the diagnosis. The MAGNIMS guidelines for space dissemination recommends the identification of T2 lesions in at least two of five locations (cortical/juxtacortical, periventricular, infratentorial, optic nerve, and spinal cord) \cite{Filippi2016}. Paraventricular region requires at least three or more lesions to be confirmed. Time dissemination recommends the presence of at least two new lesions in a follow-up MRI when compared to a baseline, or the simultaneous presence of asymptomatic lesions.

Diffusion MRI (dMRI) is found to be sensitive to tissue changes such as edema, axonal injury and demyelination, and is becoming an important MS tissue quantification strategy. A number of diffusion metrics (Mean diffusivity (MD), radial diffusivity (RD), and axial diffusivity (AD)) have become important in MS tissue quantification and analysis.  They are particularly useful when study normal-appearing white matter. Normal-appearing white matter (NAWM) is microscopically abnormal white matter tissue that is visually indistinguishable from normal white matter in T2 MRI images. They are generally defined to be at least 1 cm away from a lesions. NAWM is found to include microscopic demyelination, gliosis, small round cell infiltration, macrophages and microglial activations. Only about 28 \% of NAWM is found to be normal, and there is a 12-42 \% reduction in axonal density in these tissues. DMRI study of NAWM have found decreased fractional anisotropy and increased mean diffusivity \cite{Ciccarelli2003d,DeGroot2013}.

\subsection{Pain and MS}
Pain is a common co-morbidity of MS, with up to 50-85\% prevalence \cite{Osterberg2005}. Common pain symptoms include acute pain such as optic neuritis and painful tonic seizure, as well as chronic pain such as trigeminal neuralgia. 91\% of the patients report pain at the time of evaluation, and most experience daily pain. Both central and peripheral pain are reports in people of MS. Central pain in MS is reported to occur 27-58\% of the subjects \cite{OConnor2008}.

>>>>

\section{Multiple Sclerosis}
This chapter will review multiple sclerosis (MS) as a disorder, given the importance of MS-related TN is an independent TN subtype. MS is a chronic autoimmune demyelination disease of the central nervous system. It is a progressive disease characterized by wide-spread lesions, or plaques, within the brain, spinal cord, and optic nerve. Its pathogenesis is related to the focal inflammation and damage of the central myelin, where the disruptions of the myelin sheath negatively affect signal transduction of the central white-matter, and thereby result in severe disability. MS symptoms can include cognitive deficits, fatigue, depression, motor weakness affecting the upper and lower extremities and neuropathic pain. Currently there is no cure for MS. An estimated 2,300,000 people world wide are affected by MS, and 50\% of the patients will require motor assistance 15 years after disease onset \cite{Goldenberg2012}. The course of MS can be highly variable due to the distinct plaque distribution pattern in each individual, and therefor MS is difficult to diagnose and treat.

@@

Positive MS diagnosis is satisfied by A) At least two different lesions are present in the CNS white matter (space dissemination); B) At least two different MS exacerbation episodes must occur in the disease course (time dissemination); and C) Chronic inflammation of the CNS. One or more of these criteria result in a general diagnosis of MS. Both space and time dissemination criteria are confirmed with clinical MRI. Detailed disability ratings are confirmed by the Expanded Disability Status Scale (EDSS)\cite{Kurtzke1983}, which tracks the progression of MS symptoms in an individual.

MS is identified by the existence of focal demyelination in the CNS that manifests as lesions or plaques. Ultrastructural lesion analysis shows demyelination in the form of myelin debris, macrophages, lymphocytes and Creutzfeldt-Peters cells. Partial remyelination by oligodendrocytes can be present, as well as signs of axonal swelling and injury. In tissue samples of longstanding MS there is no presence of acute demyelination, and heavy loss of axons and myelin could be observed \cite{Filippi2012}.

@@

The MS lesions or plaques are characterized as focal hyperintensities (bright spots) in T2-weighted or proton-density images. Chronic and acute lesions could be distinguished by the presence of blood-brain-barrier (BBB) breakdown with the use of MR contrast agents such as gadolinium. Focal MS lesions commonly appear as round hyperintensities in MR images ranging from a few milimeters to centimeter in size. They are also found densely dispersed in paraventricular white matter. The presence of MR hyperintensities is not unique to MS, and therefore the location, shape and time dissemination must be considered for the diagnosis. The MAGNIMS guidelines for space dissemination recommends the identification of T2 lesions in at least two of five locations (cortical/juxtacortical, periventricular, infratentorial, optic nerve, and spinal cord) \cite{Filippi2016}. Paraventricular region requires at least three or more lesions to be confirmed. Time dissemination recommends the presence of at least two new lesions in a follow-up MRI when compared to a baseline, or the simultaneous presence of asymptomatic lesions.

Diffusion MRI (dMRI) has been found to be sensitive to tissue changes such as edema, axonal injury and demyelination, and is becoming an important MS tissue quantification strategy. A number of diffusion metrics (mean diffusivity (MD), radial diffusivity (RD), and axial diffusivity (AD)) have become important in MS tissue quantification and analysis.  They are particularly useful when study normal-appearing white matter. Normal-appearing white matter (NAWM) is white matter tissue in close proximity to MS lesions, and is visually indistinguishable from normal white matter on all conventional MRI sequences. They are generally defined to be at least 1 cm away from a lesions. NAWM is found to include microscopic demyelination, gliosis, small round cell infiltration, macrophages and microglial activations. Only about 28 \% of NAWM is found to be normal, and there is a 12-42 \% reduction in axonal density in these tissues. DMRI study of NAWM have found decreased FA and increased MD \cite{Ciccarelli2003d,DeGroot2013}.

\subsection{Pain and MS}
Pain is a common co-morbidity of MS, with up to 50-85\% prevalence \cite{Osterberg2005}. Common pain symptoms include acute pain such as optic neuritis and painful tonic seizure, as well as chronic pain such as TN. 91\% of the patients report pain at the time of evaluation, and most experience daily pain. Both central and peripheral pain are reports in people of MS. Central pain in MS is reported to occur 27-58\% of the subjects \cite{OConnor2008}.

### 2018-04-16 contents/reviews/pain.tex
Updates from ShareLaTeX

\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{descartes-reflex.JPG}
\centering
\caption{Illustration of Descartes' pain pathway from the Treatise of Man}
\label{fig:descarte}
\end{figure}

@@

The gate-control theory \cite{Melzack1965a} (\ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells that relays touch or pain sensations. The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission. The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.

@@

The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in convergent pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory. The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes the necessary information for pain. That additional processing of information in the peripheral and central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities. This entire pattern of neural activities constitute the sensory component of the pain experience. Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn (laminae V), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex. A central concept, which was also introduced by Melzacks, is the pain neuromatrix, which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}. (See section \ref{section:neuromatrix})

@@

\begin{figure}[ht]
\includegraphics[width=0.8\textwidth]{}
\centering
\caption{The axonal projections of the fast conduction ($ A\delta$) and slow conducting (C) pain stimuli in the dorsal horn.}
\label{fig:drg}
\end{figure}

The interneurons in the dorsal horn plays an important part in somatosensory integration as the sensory afferent pass from peripheral into the central nervous system. For example, the thermal grill illusion, discovered by Thunberg in 1896, involved the paradoxical feeling of heat or burning pain when adjacent skin regions are exposed to innocuous warm or cold stimuli \cite{Craig1994}. The mechanism is through the antagonistic interactions between innocuous cold A-fibers, cold heat pinch (CHP) C-fibers and warm C-fibers. Whereby the warm stimulus activates C-fibers that inhibit adjacent A-fiber activation, thereby unmasks the CHP C-fiber activations into the central nervous system.

Wide dynamic range (WDR) neurons are found primarily in the spinal dorsal horn in lamina V-VI. These neurons respond to both innocuous and painful stimuli, and are decidedly non-specific. They are found to sustain their impulse discharge during long durations of repetitive nociceptive stimulation, while nociceptive specific (NS) neurons showed a declined discharge rate within the same duration \cite{Coghill1993,Maixner1986}, thereby providing conclusive evidence that WDR neurons are sufficient for the percept of prolonged nociceptive pain.

In each spinal afferent level, the dorsal horn neurons project noceciptive signals to the brain via five distinct paths (\ref{fig:pain-pathways}). Lamina I, V-VII axons crosses the mid-line to form the \textit{spinothalamic} tract that ascend in the contralateral anterolateral column of the spinal cord, and project to thalamic nuclei. Laminae VII and VIII neurons from the \textit{spinorecticular} tract ascend via the anterolateral quadrant of the spinal cord, without crossing the midline, to reach the reticular formation and thalamus. The \textit{spinomesencephalic} tract include axons from laminae I and V, and ascend in the anterolateral quadrant to the mesencephalic reticular formation, periaqueductual gray. The \textit{cervicothalamic} tract include only the white matters of the upper two cervical segments. It takes inputs from laminae III and IV, and projects to the thalamus via the medial lemniscus. The spinohypothalamic tract contains axons from laminae I, V, and VIII, and project to hypothalamic nuclei for autonomic functions.

@@

Before reaching the neocortex, several important second order pain afferents arrive at the thalamus. The spinothalamic tracts subserving face and body sensory information terminate in the ventroposterior complex (VP, also known as ventrocaudal (Vc) nucleus), where ventroposterior-medial (VPM) nuclei and ventroposterior-lateral (VPL) nuclei are termination sites for face and body respectively. The VP complex sends third-order afferents to the primary sensory cortex (S1). Thus, the VP representation is somatotopically organized with the face region in the medial aspect, and lower limbs towards the lateral aspect. The representation is retained in the cortical projection, where the S1 somatotopy is shown to have the face region towards the lateral surface of the S1 cortical strip, and lower limb towards the superior-medial aspect.

Another termination site is the mediodorsal (MD) nucleus of the thalamus, where pain afferents of the spinothalamic tract that originated from dorsal horn lamina I layer synapse. The MD nucleus projects third order afferents to the anterior cingulate cortex (ACC), which is shown to be prominent in the emotive/suffering cognitive aspect of pain.

An important nucleus has thought to be unique in humans is the ventromedial-posterior nucleus (VMpo)\cite{Willis2002,Craig2014}. It receives afferents primarily from dorsal horn lamina I, and can be considered to be contiguous with the basal ventral medial nucleus (VMb), that forms a relay for homeostatic afferents \cite{Craig2003}. The VMPo projects to the dorsal posterior insular cortex, which is thought to involve body interoception of pain.

@@

 Important cortical regions involved in pain include anterior cingulate cortex (ACC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1). The role of these cortical regions in pain were discovered through neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}, which demonstrated consistent BOLD co-activations during painful stimulation.
Regions such as ACC, PFC and insula are also shown to be active in social pain during social exclusion and emotional stress associated with PTSD revealed changes in ACC and insula grey matter (Section \ref{fig:pain-matrix}).

@@

A recent study suggests that pain is correlated to the dynamically changing patterns of BOLD activities over time in the regions involved. Thus, the focus of the salient pain percept requires a switch away from the default mode network \cite{Kucyi2013}. The dynamic rhythms of activities that attune to and away from saliency, form a complex pain connectome \cite{Kucyi2015}. Conditions such as chronic pain therefore show a differentiated tuning of the default mode network and attention patterns of the brain due to their pathological activation of the pain matrix in the absence of detectable nociceptive stimuli \cite{Baliki2008,Legrain2009}.

\paragraph{Sensory-Discrimination}
Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the primary sensory (S1) and secondary sensory (S2) cortices. The S1 cortex is responsible for the discrimination of fine-touch sensations. S1 receives inputs directly from VP thalamus. Nonetheless, S1 has not been consistently reported in acute pain studies, possibly due to the difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within S1 \cite{bushnell1999pain}. S2, however, receives direct nociceptive input from VPI thalamus, and its BOLD activation is more consistently observed across pain fMRI studies \cite{Wager2013,Brodersen2012}. The posterior insula also receives direct nociceptive inputs from the VMPo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}. Therefore it is believe to plays an important role in pain sensory-discrimination.

\paragraph{Affective-Motivational}
The cortical regions that are involved in affective-motivational pain processing, including unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions. The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the insula receives input from the VMpo thalamus. The amygdala is important for fight-flight reflex, and regions in the prefrontal cortices are believed to be related to the cognitive aspects of pain. These regions are demonstrated to correlate with the sensory-discriminative regions during the pain experience, but can be disassociated for example in pain empathy \cite{Singer2004}. The unpleasantness of pain has been thought to be a motivational force \cite{Auvray2010}, that biases actions towards avoidance. There is evidence that the motor cortex, which is involved in action planning, and the ACC are activated during pain \cite{Erpelding2013}. The cortical excitability of the motor cortex is also shown to be altered in chronic pain \cite{Parker2016}. The motivational system is also involved in pain-reward and involves the ACC, lateral orbitofrontal cortex (OFC), and the prefrontal cortex (PFC, both ventromedial and anterior aspects) \cite{Navratilova2015}. The nucleus accumbens (NAc) BOLD activity correlated with anticipation of pain relief \cite{Baliki2010}. The reward/motivational circuitry is thought to be disrupted in chronic pain, and such disruption is related to the emotional and cognitive deficits in chronic pain sufferers.

@@

There is a recent debate pertaining the dorsal posterior insula as the "pain cortex" \cite{Segerdahl2015a}. Segerdahl et al. suggested that there is strong coupling between absolute cerebral blood flow and tonic pain intensity, and utilized arterial spin-labelling to quantify cortical activation in response to pain stimulation. The contralateral dorsal posterior insula emerged as the only cortical region with significant correlation. Tract tracing studies in monkeys has previously revealed pain-modality specific anatomy in the subregion of the dorsal posterior insula \cite{Craig2014}, and the study suggested that this is a strong supportive finding of a homologue region in humans. Critics of this finding raised issues with the low number of subjects, long sampling duration and the incompatibility of this interpretation with long standing bodies of neuroimaging literature \cite{Davis2015}.

@@

Chronic pain is pain that persists or recurrent past the healing phase following an injury \cite{Apkarian2009,Merskey2012}. It is more commonly defined as pain symptoms that persist longer than a fixed duration, most commonly 3 to 6 months, other factors to consider include psychosocial factors such as emotional and functional stress in addition to pain severity \cite{Treede2015}. The new ICD-11 revision of chronic pain classification include a new category for chronic primary pain. It is defined as "pain in one or more anatomical regions that persists or recurs for longer than 3 months, and is associated with significant emotional stress or significant functional disability" \cite{Treede2015}. It includes chronic pain conditions where the etiology is unknown. Conditions such as fibromyalgia, and irritable bowel syndrome fall under this category. Chronic pain conditions are considered to a type of maladaptive pain, whereby the pain response no longer serves its role in protecting the body from harm, but rather contributes negatively to the normal functions of life \cite{May2008}.

@@

The maladaptive mechanism responsible for neuropathic pain may manifest in any stage of the nociceptive transmission and processing circuitry. 
Noxious signal arises in the periphery, and undergoes signal transformation in the periphery and spinal ganglions, and is then subsequently processed as the pain percept in the neocortex. The pain percept is also modulated by emotional and cognitive factors, which triggers downstream modulation that ultimately affects the nociceptive signal transformation.

>>>>

\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{descartes-reflex.JPG}
\centering
\caption{Illustration of Descartes' pain pathway from the Treatise of Man \protect\cite{descartes-wikipedia}}
\label{fig:descarte}
\end{figure}

@@

The gate-control theory \cite{Melzack1965a} (Figure \ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells that relays touch or pain sensations. The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission. The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.

@@

The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory. The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes the necessary information for pain. That additional processing of information in the peripheral and central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities. This entire pattern of neural activities constitute the sensory component of the pain experience. Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn (laminae V), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex. A central concept, which was also introduced by Melzack, is the pain neuromatrix, which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}. (See section \ref{section:neuromatrix})

@@

\begin{figure}[ht]
\includegraphics[width=0.8\textwidth]{dorsal-lamina.pdf}
\centering
\caption{The axonal projections of the fast conduction ($ A\delta$) and slow conducting (C) pain stimuli in the dorsal horn.}
\label{fig:drg}
\end{figure}

The interneurons in the dorsal horn plays an important part in somatosensory integration as the sensory afferent pass from peripheral into the central nervous system. For example, the wide dynamic range (WDR) neurons are found primarily in the spinal dorsal horn in lamina V-VI. These neurons respond to both innocuous and noxious stimuli, and are decidedly non-specific. They are found to sustain their impulse discharge during long durations of repetitive nociceptive stimulation, while nociceptive specific (NS) neurons showed a declined discharge rate within the same duration \cite{Coghill1993,Maixner1986}, thereby providing evidence that WDR neurons are responsive to prolonged nociceptive pain.

In each spinal afferent level, the dorsal horn neurons project noceciptive signals to the brain via five distinct paths (Figure \ref{fig:pain-pathways}). Lamina I, V-VII axons crosses the mid-line to form the \textit{spinothalamic} tract that ascend in the contralateral anterolateral column of the spinal cord, and project to thalamic nuclei. Laminae VII and VIII neurons from the \textit{spinorecticular} tract ascend via the anterolateral quadrant of the spinal cord, without crossing the midline, to reach the reticular formation and thalamus. The \textit{spinomesencephalic} tract include axons from laminae I and V, and ascend in the anterolateral quadrant to the mesencephalic reticular formation, including the periaqueductual gray. The \textit{cervicothalamic} tract include only the white matters of the upper two cervical segments. It takes inputs from laminae III and IV, and projects to the thalamus via the medial lemniscus. The spinohypothalamic tract contains axons from laminae I, V, and VIII, and project to hypothalamic nuclei for autonomic functions.

@@

Before reaching the neocortex, several important second order pain afferents arrive at the thalamus. The spinothalamic tracts subserving face and body sensory information terminate in several regions of the thalamus. A major termination site is the ventroposterior complex (VP, also known as ventrocaudal (Vc) nucleus), where ventroposterior-medial (VPM) nuclei and ventroposterior-lateral (VPL) nuclei are termination sites for the face and body respectively. The VP complex sends third-order afferents to the primary sensory cortex (S1). Thus, the VP representation is somatotopically organized with the face region in the medial aspect, and lower limbs towards the lateral aspect. The representation is retained in the cortical projection, where the S1 somatotopy is shown to have the face region towards the lateral surface of the S1 cortical strip, and lower limb towards the superior-medial aspect.

Another spinothalamic termination site is the mediodorsal (MD) nucleus of the thalamus, where pain afferents of the spinothalamic tract that originated from dorsal horn lamina I layer synapse. The MD nucleus projects third order afferents to the anterior cingulate cortex (ACC), which is shown to be prominent in the emotive/suffering cognitive aspect of pain.

An important nucleus thought to be unique in humans is the ventromedial-posterior nucleus (VMpo)\cite{Willis2002,Craig2014}. It receives inputs primarily from dorsal horn lamina I, and can be considered to be contiguous with the basal ventral medial nucleus (VMb), that forms a relay for homeostatic afferents \cite{Craig2003}. The VMPo projects to the dorsal posterior insular cortex, which is thought to involve body interoception of pain.

@@

 Important cortical regions involved in pain include anterior cingulate cortex (ACC), insula, prefrontal cortex (PFC), thalamus, periaqueductal grey (PAG), secondary somatosensory cortex (SII) and motor cortex (M1). The role of these cortical regions in pain were studied in neuroimaging fMRI studies \cite{Apkarian2013c,Wager2013,Davis2012a}, which demonstrated consistent BOLD co-activations during noxious stimulation. (Figure \ref{fig:pain-matrix}).

@@

A recent study suggests that pain is correlated to the dynamically changing patterns of BOLD activities over time in the regions involved. Thus, the focus of the salient pain percept requires a switch away from the default mode network \cite{Kucyi2013}. The dynamic rhythms of activities that attune to and away from saliency, form a dynamic pain connectome \cite{Kucyi2015}. Chronic pain therefore demonstrated differential tuning of the default mode network and attention patterns of the brain due to their pathological activation of the pain matrix in the absence of detectable noxious stimuli \cite{Baliki2008,Legrain2009}.

\paragraph{Sensory-Discrimination}
Cortical regions associated with sensory-discriminative aspect of pain, such as the location and intensity of pain, include the S1 ana S2 cortices. The S1 cortex is responsible for the discrimination of fine-touch sensations. S1 receives inputs directly from VP thalamus. Nonetheless, S1 has not been consistently reported in acute pain studies, possibly due to the difficulty in detecting pain-related activations as there are relatively small number of nociceptive neurons within S1 \cite{bushnell1999pain}. S2, however, receives direct nociceptive input from VPI thalamus, and its BOLD activation is more consistently observed across pain fMRI studies \cite{Wager2013,Brodersen2012}. The posterior insula also receives direct nociceptive inputs from the VMpo thalamus \cite{Craig2003}, and shows strong functional and structural connectivity with S2 \cite{Wiech2014}. Therefore the insula, as well as S1, are believed to plays an important role in sensory-discrimination aspect of pain.

\paragraph{Affective-Motivational}
The cortical regions that are involved in affective-motivational pain processing, including unpleasantness and flight response, involve both regions that receive direct projection from the medial thalamus, as well as indirection regions. The anterior cingulate cortex (ACC) receives direct cortical projection from the medial dorsal nucleus, while the anterior insula receives input from the VMpo thalamus. The amygdala is important for fight-flight reflex, and regions in the prefrontal cortices are believed to be related to the cognitive aspects of pain. These regions are demonstrated to correlate with the sensory-discriminative regions during the pain experience, but can be disassociated for example in pain empathy \cite{Singer2004}. The unpleasantness of pain has been thought to be a motivational force \cite{Auvray2010}, that biases actions towards avoidance. There is evidence that the motor cortex, which is involved in action planning, and the ACC are activated during pain \cite{Erpelding2013}. The cortical excitability of the motor cortex is also shown to be altered in chronic pain \cite{Parker2016}. The motivational system is also involved in pain-reward and involves the ACC, lateral orbitofrontal cortex (OFC), and the prefrontal cortex (PFC, both ventromedial and anterior aspects) \cite{Navratilova2015}. The nucleus accumbens (NAc) BOLD activity correlated with anticipation of pain relief \cite{Baliki2010}. The reward/motivational circuitry is thought to be disrupted in chronic pain, and such disruption is related to the emotional and cognitive deficits in chronic pain sufferers.

@@

There is a recent debate pertaining the dorsal posterior insula as the "pain cortex" \cite{Segerdahl2015a}. Segerdahl et al. suggested that there is strong coupling between absolute cerebral blood flow and tonic pain intensity, and utilized arterial spin-labelling to quantify cortical activation in response to pain stimulation. The contralateral dorsal posterior insula emerged as the only cortical region with significant correlation. Tract tracing studies in monkeys has previously revealed pain-modality specific anatomy in the subregion of the dorsal posterior insula \cite{Craig2014}, and the study suggested that this is a strong supportive finding of a homologue region in humans. Critics of this finding raised issues with the low number of subjects, long sampling duration and the incompatibility of this interpretation with long standing bodies of neuroimaging and neurophysiological literature \cite{Davis2015}.

@@

Chronic pain is pain that persists past the healing phase following an injury \cite{Apkarian2009,Merskey2012}. It is more commonly defined as pain symptoms that persist longer than a fixed duration, most commonly 3 to 6 months. Other factors to consider include psychosocial factors such as emotional and functional stress in addition to pain severity \cite{Treede2015}. The new ICD-11 revision of chronic pain classification include a new category for chronic primary pain. It is defined as "pain in one or more anatomical regions that persists or recurs for longer than 3 months, and is associated with significant emotional stress or significant functional disability" \cite{Treede2015}. It includes chronic pain conditions where the etiology is unknown. Conditions such as fibromyalgia, and irritable bowel syndrome fall under this category. Chronic pain conditions are considered to a type of maladaptive pain, whereby the pain response no longer serves its role in protecting the body from harm, but rather contributes negatively to the normal functions of life \cite{May2008}.

@@

The maladaptive mechanism responsible for neuropathic pain may manifest in any stage of the nociceptive transmission, processing, and modulatory circuitry. 
Noxious signal arises in the periphery, and undergoes signal transformation in the periphery and spinal ganglions, and is then subsequently processed as the pain percept in the neocortex. The pain percept is also modulated by emotional and cognitive factors, which triggers downstream modulation that ultimately affects the nociceptive signal transformation.
